### 2013-10-15 Publications/2013-TMI/04-methods.tex
Finished inclusion of @zosso comments
Some subscripts and mathematical symbols must be reviewed
because they are not consistent.

\subsection{Proposed model}
\label{sec:methods_map}
In this paper we formulate the joint registration-segmentation
problem as follows. We look for a spatial mapping (with, typically,
$n=3$ for the 3-dimensional case):
\begin{align*}
U\colon T \subset \mathbb{R}^n &\to D \subset \mathbb{R}^n \\
\mathbf{x} &\mapsto \mathbf{x}' =\mathbf{x}+u(\mathbf{x}),
\end{align*}
such that the known contours in anatomical space $T$ optimally segment
the diffusion space $D$.

@@

Further, within each region $\Omega_k$ defined by the contours $\Gamma$
trough duality, we assume the features to be i.i.d.:
\begin{equation}
P\left( F(\mathbf{x}') \mid U,\Gamma_{ij} \right) = p_k( F(\mathbf{x}')) \quad
\forall \mathbf{x} \in \Omega_k.
\label{eq:likelihood}
\end{equation}

For convenience, and because this has been shown to be an appropriate
approximation \citep{bishop_pattern_2006}, each such region is modeled
by a multivariate normal distribution with region descriptors
$\Theta_k = \lbrace \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_{k} \rbrace$:

\begin{equation}
p_k( F(\mathbf{x}') ) = \mathcal{N} ( f \mid \Theta_k ).
\label{eq:multivariate_normal}
\end{equation}

\todo{I modified this slightly}
Thus, defining the feature observed as $\mathbf{f}'=F(\mathbf{x}')$ and 
the squared \emph{Mahalanobis distance} of $\mathbf{f}'$ with respect
of the descriptors of region $k$ as
$\Delta^2_k (\mathbf{f}' \mid \Theta_k ) = (\mathbf{f}' - \boldsymbol{\mu}_k)^T \, \boldsymbol{\Sigma}^{-1}_k \, (\mathbf{f}' - \boldsymbol{\mu}_k)$,
we write:
\begin{align}
P( F \mid U, \Omega ) &= \underset{k}{\prod} \underset{\mathbf{x} \in \Omega_k}{\prod}
\mathcal{N} ( f \mid \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_{k} ) \notag\\
&= \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{k}\right|}}\,{e^{\left(-\frac{1}{2}  \Delta^2_l (\mathbf{f}' \mid \Theta_k ) \right)}}.
\label{eq:pdf}
\end{align}

@@

\begin{equation*}
P(U) = \underset{\mathbf{x}}{\prod} p(u)
\end{equation*}
with further:
\begin{align*}
p(u) &= p_0(u) \, p_1(u) \notag\\
p_0(u) &= \mathcal{N}( u \mid 0, A^{-1}) \notag\\
p_1(u) &= \mathcal{N}( Du \mid 0, B^{-1}),
\end{align*}
that consider that the distortion and its gradient have zero
mean and variance governed by $A$ and $B$. Since the anisotropy
is technically aligned with the imaging axes, these can
be simplified:
\begin{align}
p_0(u) &= \underset{\mathbf{x}}{\sum} \mathcal{N}( u(\mathbf{x}) \mid 0, \frac{1}{\sqrt{\alpha_{x,y,z}}}) \notag\\
p_1(u) &= \underset{\mathbf{x}}{\sum} \mathcal{N}( \nabla_{x,y,z}u(\mathbf{x}) \mid 0, \frac{1}{\sqrt{\beta_{x,y,z}}})
\label{eq:priors}
\end{align}

Finally, we can turn the \gls{map} problem into 
a variational one
applying the following log-transform:
\begin{align}
E(F \mid U) &= -\log \underset{k}{\prod}
\underset{\mathbf{x} \in \Omega_k}{\prod} 
p_k( \mathbf{f}')\,p_0( u(\mathbf{x}))\,p_1( u(\mathbf{x})) \notag\\
&= C + \underset{k}{\sum}
\underset{\mathbf{x} \in \Omega_k}{\sum}
\Delta^2_k (\mathbf{f}' \mid \Theta_k ) \notag\\
&+ \underset{d=\{x,y,z\}}{\sum} \alpha_d u^2_d
+ \underset{d=\{x,y,z\}}{\sum} \beta_d \|\nabla u_d\|^2,
\label{eq:energy}
\end{align}
that is the dual expression to the energy functional corresponding
to a discrete \gls*{acwe} framework \citep{chan_active_2001}
with anisotropic regularization as studied in
\citep{nagel_investigation_1986}.

\subsection{Numerical Implementation}
\label{sec:numerical_implementation}
%
Let us denote $\{c_i\}_{i=1 \ldots N_c}$ the nodes of one or several shape-prior
surface(s). In our application, precise tissue interfaces of interest 
extracted from a high-resolution, anatomically correct reference volume. 
On the other hand, we have a number of \gls{dwi}-derived features at each
voxel of the volume. Let us denote by $\mathbf{x}$ the voxel and $f(\mathbf{x}) = 
[ f_1, f_2, \ldots, f_N]^T(\mathbf{x})$ its associated feature vector.

\subsubsection{Deformation field}
\label{sec:deformation_field}
The transformation from reference into \gls{dwi} coordinate space is 
achieved through a dense deformation field $u(\mathbf{x})$, such that:
\begin{equation}
c_i' = T\{c_i\} = c_i + u(c_i),
\end{equation}
what is equivalent to \eqref{eq:omega_tf}. Since the nodes of the anatomical 
surfaces might lay off-grid, it is required to derive $u(\mathbf{x})$ from a discrete 
set of parameters $\{u_k\}_{k=1 \ldots K}$. Densification is achieved through 
a set of associated basis functions $\psi_k$ (e.g. rbf, interpolation splines):
%
\begin{equation}
u(\mathbf{x}) = \sum_k \psi_k(\mathbf{x}) u_k
\end{equation}
%
In our implementation, $\psi_k$ is chosen to be a tensor-product B-Spline kernel.
Then, the transformation writes
%
\begin{equation}
\label{eq:transformation}
c_i' = T\{c_i\} = c_i + u(c_i) = c_i + \sum_k \psi_k(c_i)\,u_k
\end{equation}

Based on the current estimate of the distortion $u$, we can compute 
``expected samples'' within the shape prior projected into the \gls{dwi}.
Thus, we now estimate region descriptors of the \gls{dwi} features 
$f(\mathbf{x})$ of the regions defined by the priors in \gls{dwi} space.
%
Using the region descriptors derived in \autoref{sec:methods_map}, we propose
an \gls{adf}-like, \gls{acwe}-based, piece-wise constant, image segmentation
model (where the unknown is the deformation field)
\cite{chan_active_2001} with the energy functional obtained in 
\eqref{eq:final_map_energy}. This inverse problem is ill-posed
\cite{bertero_ill-posed_1988,hadamard_sur_1902}.
In order to account for deformation field regularity and to render the 
problem well-posed, we include limiting and regularization terms into 
the energy functional \cite{morozov_linear_1975,tichonov_solution_1963}:
%
\begin{multline}
E(u) = \sum\limits_l \int_{\Omega'_l} \Delta^2_l (\mathbf{f}) \,d\mathbf{x} \\
+ \int u^T \, A \, u \, d\mathbf{x} + \int \tr\{(\nabla u^T)^T B (\nabla u^T)\} d\mathbf{x}
\label{eq:complete_energy}
\end{multline}
%
These regularity terms ensure that the segmenting contours in 
\gls{dwi} space are still close to their native shape. The model
easily allows to incorporate inhomogeneous and anisotropic 
regularization \cite{nagel_investigation_1986} to better regularize
the \gls{epi} distortion.

\subsubsection{Shape gradients}
To compute the gradient-descent of the data-term domain integrals
in \eqref{eq:complete_energy} with respect to the underlying 
deformation field, we use shape gradients 
\cite{jehan-besson_dream2s:_2003,herbulot_segmentation_2006}.

Let $\Omega$ be an image domain and $d\Omega$ its boundary. Further, 
$r(\mathbf{x})$ is an ``arbitrary'' function over the image domain. 
We now derive the domain integral w.r.t. the contour evolution 
parameter $t$ (time):
\begin{equation}
\frac{\partial}{\partial t} \int_\Omega r(\mathbf{x}) d\mathbf{x} = \int_\Omega \frac{\partial r}{\partial t}(\mathbf{x}) d\mathbf{x} - \int_{d\Omega} r(\mathbf{x}) \left\langle \frac{\partial{d\Omega}}{\partial t}, N_{d\Omega}\right\rangle d\mathbf{x}
\label{eq:shape_gradients}
\end{equation}
where $\left\langle\frac{\partial{d\Omega}}{\partial t}, N_{d\Omega}\right\rangle$ is 
the projection of the boundary movement on the unit inward normal. We recall
here that equation \eqref{eq:shape_gradients} stands the bridge between the 
explicit level set formulations surveyed in \autoref{sec:methods_background}
by updating the distortion along the steepest energy descent:
\begin{equation}
\frac{\partial u(\mathbf{x},t)}{\partial t} = - \frac{\partial E(u)}{\partial u},
\end{equation}
what closes the loop with the general expression of the \glspl{adf} introduced in
\eqref{eq:active_deformation_fields}.

\subsubsection{Operator splitting}
In order to make the energy minimization computationally more tractable, 
we propose the following operator splitting: Let us optimize the data terms 
and the regularity terms on separate copies of the deformation field, 
now called $u$ and $v$, constrained to be equal:
\begin{multline}
E(u,v) = \sum\limits_l \int_{\Omega'_l(u)} \Delta^2_l (\mathbf{f}) \,d\mathbf{x} \\
+ \int v^T \, A \, v \, d\mathbf{x} + \int \tr\{(\nabla v^T)^T B (\nabla v^T)\} d\mathbf{x}
\label{eq:operator_splitting}
\end{multline}
and now
\begin{equation}
\min_{u,v} \{ E \} \quad s.t. \quad u = v.
\end{equation}

In order to take the equality constraint into account, we may make use of 
augmented Lagrangians (a combination of Lagrangian multipliers and penalty 
terms on the constraint) \cite{bertsekas_multiplier_1976,
glowinski_augmented_1989,nocedal_numerical_2006}:
\begin{equation}
AL(u,v,\lambda,r) = E(u,v) + \langle \lambda, u-v \rangle + \frac{r}{2} \| u - v \|_2^2
\end{equation}

To solve the constraint minimization problem, we may now optimize the 
Augmented Lagrangian in an iterative way:
\begin{equation}
\left\lbrace 
\begin{array}{rcl}
u^{t+1} &=& \argmin_{u} AL(u,v^t,\lambda^t,r)\\
v^{t+1} &=& \argmin_{v} AL(u^{t+1},v,\lambda^t,r)\\
\lambda^{t+1} &=& \lambda^t + \rho(u^{t+1}-v^{t+1})
\end{array}\right.
\end{equation}
where typically $0 < \rho < r$. The two sub-minimization problems will 
now be much easier to handle than the original complete problem.

\subsubsection{Minimization w.r.t. $u$}
The first minimization problem optimizes the data-term. The problem is equivalent 
to minimizing the following energy:
\begin{multline}
E(u,v) = \sum\limits_l \int_{\Omega'_l(u)} \Delta^2_l (\mathbf{f}) \,d\mathbf{x}
+ \langle \lambda, u-v \rangle + \frac{r}{2} \| u - v \|_2^2
\end{multline}
where we identify one instance of domain integrals of the form $\int_\Omega r(\mathbf{x}) 
d\mathbf{x}$ for each region $l$. Optimality requires the derivative of this energy 
with respect to the parameters $u$ to be null. At this point, we decide to ignore 
the influence of the boundary shift on the statistics of the regions (i.e. moving the 
boundary does not significantly impact the $\mu$ and $\Sigma$ descriptors). This means 
that we can drop the derivative of $r(\mathbf{x})$ w.r.t. contour evolution. 
What remains are the surface integrals at the subdomain interfaces, plus the Lagrangian 
and penalty terms:

\begin{multline}
\frac{\partial E}{\partial u_k} = \sum\limits_l \int_{d\Omega'_l} \left[ \Delta^2_{O(l)} (\mathbf{f}) - \Delta^2_l (\mathbf{f}) \right]
\left\langle\frac{\partial{d\Omega'_l}}{\partial t}, N_{d\Omega'_l}\right\rangle \,d\mathbf{x} \\
+ \lambda_k + r(u_k - v_k) = 0
\label{eq:energy_gradient}
\end{multline}
where $O(l)$ is a function that returns the neighboring region at $\mathbf{x}$ (i.e.
the region \emph{outside} the current location in the contour). For the shake of
simple notation, we will refer as $c_i$ to the positions 
of points belonging to any of the existing contours $d\Omega_l$ (with 
$i \in [1 \ldots N_c]$ and $N_c$ the total number of control points).
Given the deformation field interpolation stated in \autoref{sec:deformation_field}, 
the boundary moves according to
\begin{equation}
\frac{\partial c_i}{\partial u_k} = \psi_k(c_i)\,e_a
\end{equation}
where $e_a$ is the unit vector along direction $a \in \mathbb{R}^n$ and $n$ the dimension
of the image. Thus, if we name the inward unit normal of the corresponding transformed 
contour at position $c'_i$ as $\hat{\mathbf{n}}_{c'_i}$,
\begin{equation}
\left\langle\frac{\partial c_i}{\partial u_k}, N_{c_i}\right\rangle = \psi_k(c_i)\,\hat{\mathbf{n}}_{c'_i}.
\end{equation}
The discrete implementation of \eqref{eq:energy_gradient} is straightforward as
we have a triangularized mesh representation of the interfacing surfaces:
\begin{multline}
\frac{\partial E}{\partial u_k} = \sum\limits_{N_c} \left[ \Delta^2_{O(l)} (\mathbf{f}) - \Delta^2_l (\mathbf{f}) \right]
\, \psi_{c_i}(\mathbf{x}_k)\,\hat{\mathbf{n}}_{c'_i} \\
+ \lambda_k + r(u_k - v_k) = 0,
\label{eq:discrete_energy_gradient}
\end{multline}
where we have swapped $\psi_k(c_i)$ by $\psi_{c_i}(\mathbf{x}_k)$ for
convenience in implementation, taking the advantage that they are
equivalent. Subsequently, the optimal distortion $u_k$ 
is found at each iteration as:
\begin{multline}
u_k^{t+1} = v_k^t - \frac{1}{r}\lambda_k^{t} \\
- \frac{1}{r} \sum\limits_{N_c} \left[ \Delta^2_{O(l)} (\mathbf{f}) - \Delta^2_l (\mathbf{f}) \right]
\, \psi_{c_i}(\mathbf{x}_k)\,\hat{\mathbf{n}}_{c'_i}
\label{eq:update_u}
\end{multline}

\subsubsection{Minimization w.r.t. $v$}
For the optimization w.r.t. $v$, the relevant energy writes
\begin{align}
E(v) &= \int v^T A v d\mathbf{x} + \int \tr\{(\nabla v^T)^T B (\nabla v^T)\} d\mathbf{x}\\
&\quad + \langle \lambda, u-v \rangle + \frac{r}{2} \| u - v \|_2^2\nonumber
\end{align}
The associated Euler-Lagrange equation is found as:
\begin{equation}
2 ( A v - B \Delta v ) + rv = ru + \lambda
\end{equation}
which easily translates into Fourier domain:
\begin{equation}
v^{t+1} = \mathcal{F}^{-1}\left\{ \frac{\mathcal{F}\{ru + \lambda\}}{\mathcal{F}\{( 2 A+r)\mathcal{I}-2 B \Delta\}} \right\}
\end{equation}
where $\mathcal{I}$ denotes the identity operator.

@@

\subsubsection{Lagrangian multiplier update}
At each iteration, the Lagrangian multipliers are updated as noted before:
\begin{equation}
\lambda^{t+1} = \lambda^t + \rho(u^{t+1}-v^{t+1})
\end{equation}

@@

\subsubsection{Convergence}
In order to ``fixate'' the evolution when close to convergence,
it is advised to slightly increase the penalty weight $r$ at each 
iteration. Note that as can be seen in the above equations, $r$ 
governs the step-size or leash-length at each iteration, i.e. the 
amount by which the new estimate $u$ may move away from the preceding 
$v$ and vice-versa.

\subsubsection{Efficient field densification}
As long as the dense deformation field is iteratively interpolated
from the same set of control points that define the $L-1$ contours,
it is possible to pre-cache all the $\psi_k(c_i)$ weights into a
sparse matrix for fast densification. As well, we achieve a 
diffeomorphic transform by ... 
\todo[inline]{this is actually false (IMHO), so far :(. We will need
to add some more conditions to the interopolating kernel so that
the original values of the transform in $c_i$ can be recovered from 
interpolated values of $u_k$}

>>>>

\subsection{Proposed model}
\label{sec:methods_map}
In this paper we formulate the joint registration-segmentation
problem as follows. We look for a spatial mapping (with, typically,
$n=3$ for the 3-dimensional case):
\begin{align}
U\colon T \subset \mathbb{R}^n &\to D \subset \mathbb{R}^n \notag\\
\mathbf{x} &\mapsto \mathbf{x}' =\mathbf{x}+u(\mathbf{x}),
\label{eq:deformation_field}
\end{align}
such that the known contours in anatomical space $T$ optimally segment
the diffusion space $D$.

@@

Further, within each region $\Omega_l$ defined by the contours $\Gamma$
trough duality, we assume the features to be i.i.d.:
\begin{equation}
P\left( F(\mathbf{x}') \mid U,\Gamma_{ij} \right) = p_l( F(\mathbf{x}')) \quad
\forall \mathbf{x} \in \Omega_l.
\label{eq:likelihood}
\end{equation}

For convenience, and because this has been shown to be an appropriate
approximation \citep{bishop_pattern_2006}, each such region is modeled
by a multivariate normal distribution with region descriptors
$\Theta_l = \lbrace \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} \rbrace$:

\begin{equation}
p_l( F(\mathbf{x}') ) = \mathcal{N} ( f \mid \Theta_l ).
\label{eq:multivariate_normal}
\end{equation}

\todo{I modified this slightly}
Thus, defining the feature observed as $\mathbf{f}'=F(\mathbf{x}')$ and 
the squared \emph{Mahalanobis distance} of $\mathbf{f}'$ with respect
of the descriptors of region $l$ as
$\Delta^2_l (\mathbf{f}' \mid \Theta_l ) = (\mathbf{f}' - \boldsymbol{\mu}_l)^T \, \boldsymbol{\Sigma}^{-1}_l \, (\mathbf{f}' - \boldsymbol{\mu}_l)$,
we write:
\begin{align}
P( F \mid U, \Omega ) &= \underset{l}{\prod} \underset{\mathbf{x} \in \Omega_l}{\prod}
\mathcal{N} ( f \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} ) \notag\\
&= \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}  \Delta^2_l (\mathbf{f}' \mid \Theta_l ) \right)}}.
\label{eq:pdf}
\end{align}

@@

\begin{equation*}
P(U) = \underset{\mathbf{x}}{\prod} p(u)
\end{equation*}
with further:
\begin{align*}
p(u) &= p_0(u) \, p_1(u) \notag\\
p_0(u) &= \mathcal{N}( u \mid 0, A^{-1}) \notag\\
p_1(u) &= \mathcal{N}( Du \mid 0, B^{-1}),
\end{align*}
that consider that the distortion and its gradient have zero
mean and variance governed by $A$ and $B$. Since the anisotropy
is technically aligned with the imaging axes, these can
be simplified:
\begin{align}
p_0(u) &= \underset{\mathbf{x}}{\sum} \mathcal{N}( u(\mathbf{x}) \mid 0, \frac{1}{\sqrt{a_{x,y,z}}}) \notag\\
p_1(u) &= \underset{\mathbf{x}}{\sum} \mathcal{N}( \nabla_{x,y,z}u(\mathbf{x}) \mid 0, \frac{1}{\sqrt{b_{x,y,z}}})
\label{eq:priors}
\end{align}

Finally, we can turn the \gls{map} problem into 
a variational one
applying the following log-transform:
\begin{align}
E(F \mid U) &= -\log \underset{l}{\prod}
\underset{\mathbf{x} \in \Omega_l}{\prod} 
p_l( \mathbf{f}')\,p_0( u(\mathbf{x}))\,p_1( u(\mathbf{x})) \notag\\
&= C + \underset{l}{\sum}
\underset{\mathbf{x} \in \Omega_l}{\sum}
\Delta^2_l (\mathbf{f}' \mid \Theta_l ) \notag\\
&+ \underset{x,y,z}{\sum} a_d u^2_d
+ \underset{x,y,z}{\sum} b_d \|\nabla u_d\|^2,
\label{eq:energy}
\end{align}
that is the dual expression to the energy functional corresponding
to a discrete \gls*{acwe} framework \citep{chan_active_2001}
with anisotropic regularization as studied in
\citep{nagel_investigation_1986}.

{\color{red} {Using the region descriptors derived in \autoref{sec:methods_map}, we propose
an \gls{adf}-like, \gls{acwe}-based, piece-wise constant, image segmentation
model (where the unknown is the deformation field)
\cite{chan_active_2001} with the energy functional obtained in 
\eqref{eq:final_map_energy}. This inverse problem is ill-posed
\cite{bertero_ill-posed_1988,hadamard_sur_1902}.
In order to account for deformation field regularity and to render the 
problem well-posed, we include limiting and regularization terms into 
the energy functional \cite{morozov_linear_1975,tichonov_solution_1963}:
%
\begin{multline}
E(u) = \sum\limits_l \int_{\Omega'_l} \Delta^2_l (\mathbf{f}) \,d\mathbf{x} \\
+ \int u^T \, A \, u \, d\mathbf{x} + \int \tr\{(\nabla u^T)^T B (\nabla u^T)\} d\mathbf{x}
\label{eq:complete_energy}
\end{multline}
%
These regularity terms ensure that the segmenting contours in 
\gls{dwi} space are still close to their native shape. The model
easily allows to incorporate inhomogeneous and anisotropic 
regularization \cite{nagel_investigation_1986} to better regularize
the \gls{epi} distortion.}}

\todo[inline]{this last paragraphs in red color were located a bit
further, but I think we can keep some of the refs and still reproduce
the continuous version of the energy functional}

\subsection{Numerical Implementation}
\label{sec:numerical_implementation}
%
Let us denote $\{c_i\}_{i=1 \ldots N_c}$ the nodes of one or several shape-prior
surface(s). In our application, precise tissue interfaces of interest 
extracted from a high-resolution, anatomically correct reference volume. 
On the other hand, we have a number of \gls{dwi}-derived features at each
voxel of the volume. Let us denote by $\mathbf{x}$ the voxel and $F(\mathbf{x}) = 
[ f_1, f_2, \ldots, f_N]^T(\mathbf{x})$ its associated feature vector.

\subsubsection{Deformation field}
\label{sec:deformation_field}
The transformation from ``T1-'' into ``\gls{dwi}-'' coordinate space is 
achieved through a dense deformation field $u(\mathbf{x})$, such that:
\begin{equation}
c_i' = U\{c_i\} = c_i + u(c_i),
\label{eq:nodes_tfm}
\end{equation}
where $U$ is defined in \eqref{eq:deformation_field}. Since the nodes of the anatomical 
surfaces might lay off-grid, it is required to derive $u(\mathbf{x})$ from a discrete 
set of parameters $\{u_k\}_{k=1 \ldots K}$. Densification is achieved through 
a set of associated basis functions $\psi_k$:
%
\begin{equation}
u(\mathbf{x}) = \sum_k \psi_k(\mathbf{x}) u_k
\label{eq:intp_kernel}
\end{equation}
%
In our implementation, $\psi_k$ is chosen to be a tensor-product B-Spline kernel
of degree 3 ($B_3$).
Then, introducing \eqref{eq:intp_kernel} into \eqref{eq:nodes_tfm} and replacing
$\psi$ by the actual kernel function, the transformation writes
%
\begin{equation}
c_i' = c_i + \sum_k \left[ u_k \, \underset{x,y,x}{\prod} B_3(c_{i,d} - \mathbf{x}_{k,d}) \right].
\label{eq:transformation}
\end{equation}

Based on the current estimate of the distortion $u$, we can compute 
``expected samples'' within the shape prior projected into the \gls{dwi}.
Thus, we now estimate region descriptors of the \gls{dwi} features 
$\mathbf{f}'$ of the regions defined by the priors in \gls{dwi} space.

\subsubsection{Gradient descent}
To find the minimum of the energy functional \eqref{eq:energy},
we propose a gradient-descent approach with respect to the underlying 
deformation field through the following partial differential equation:
\begin{equation}
\frac{\partial u(\mathbf{x},t)}{\partial t} \propto - \frac{\partial E(u)}{\partial u},
\label{eq:general_gradient_descent}
\end{equation}
with $t$ being an artificial time parameter of the contour
evolution, included to express the iterative character of the approach.
Now, we can introduce \eqref{eq:energy} in
\eqref{eq:general_gradient_descent}:
\begin{align}
\frac{\partial E(u)}{\partial u^d_k} &=
\frac{ \partial }{\partial u^d_k} \lbrace
C + \underset{l}{\sum}
\underset{\mathbf{x} \in \Omega_l}{\sum}
\Delta^2_l (\mathbf{f}' \mid \Theta_l ) \notag\\
&+ \underset{a}{\sum} a^d ({u^d})^2
+ \underset{a}{\sum} b^d \|\nabla u^d\|^2
\rbrace
\label{eq:gradient_descent}
\end{align}

Whereas related \glspl*{adf} introduced in \autoref{sec:methods_background}
make use of explicit level-set formulations to solve \eqref{eq:gradient_descent},
we alternatively use \emph{shape-gradients}
\cite{jehan-besson_dream2s:_2003,herbulot_segmentation_2006}.
Let $r(\mathbf{x})$ be an ``arbitrary'' function over the image domain,
and $\Omega$ a certain image region with $\Gamma_{ij}$ its corresponding
outer boundary as defined in \autoref{sec:methods_map}.
We now derive the domain integral w.r.t. $t$:
\begin{equation*}
\frac{\partial}{\partial t} \int_\Omega r(\mathbf{x}') d\mathbf{x} = \int_\Omega \frac{\partial}{\partial t}r(\mathbf{x}') d\mathbf{x} - \int_{\Gamma_{ij}} r(\mathbf{x}') \left\langle \frac{\partial \Gamma_{ij} }{\partial t}, N_{\Gamma_{ij}}\right\rangle d\mathbf{x}
\end{equation*}
where $\left\langle\frac{\partial\Gamma_{ij}}{\partial t}, N_{\Gamma_{ij}}\right\rangle$ is 
the projection of the boundary movement on the unit inward normal $N_{\Gamma_{ij}}$. Assuming
that the region descriptors $\Theta_l$ vary slowly enough, we can consider
that $\frac{\partial}{\partial t} r(\mathbf{x}') = 0$ and thus:
\begin{equation}
\frac{\partial}{\partial t} \int_\Omega r(\mathbf{x}') d\mathbf{x} = 
- \int_{\Gamma_{ij}} r(\mathbf{x}') \left\langle \frac{\partial \Gamma_{ij} }{\partial t}, N_{\Gamma_{ij}}\right\rangle d\mathbf{x}
\label{eq:shape_gradients}
\end{equation}

Therefore, we can apply a discretized interpretation of \eqref{eq:shape_gradients}
to compute the data term in \eqref{eq:gradient_descent} as follows:
\begin{align}
\frac{\partial E_{data}(u)}{\partial u^d_k} &=
\underset{l}{\sum} \frac{ \partial }{\partial u^d_k} \lbrace
 \underset{\mathbf{x} \in \Omega_l}{\sum}
\Delta^2_l (\mathbf{f}') \rbrace \notag\\
&= \underset{ij}{\sum} \underset{\mathbf{c}' \in \Gamma_{ij}}{\sum}
\left[ \Delta^2_i (\mathbf{f}' ) - \Delta^2_j (\mathbf{f}')\right]
\left\langle \frac{\partial \mathbf{c}'}{\partial {u^a_k}'}, \mathbf{n_i}'\right\rangle,
\label{eq:gradient_wshape}
\end{align}
in this case, the formulation has been adapted to the non-binary case, $ij$
being any pair of neighboring regions, and $\Gamma_{ij}$ the contour separating
them such that $\mathbf{x}' = \mathbf{c}' \in\Gamma_{ij} \iff \mathbf{x}\in \partial\Omega_i \cap \partial\Omega_j$
and $\mathbf{n_i}'$ is the unit inward normal to the contour at $c_i'$.

Finally, we can compute:
\begin{align}
\frac{\partial}{\partial {u^a_k}'}{\mathbf{c}_i}' &= \frac{\partial}{\partial {u^a_k}'} 
\lbrace c_i + \sum_k \psi_k(\mathbf{c_i}) u_k \rbrace
= \psi_k(\mathbf{c_i})\, \hat{\mathbf{e}}^a
\end{align}
where $\hat{\mathbf{e}}^a$ is the unit vector along direction $a$. Projecting
this gradient onto the surface normal,
$\left\langle \frac{\partial \mathbf{c}'}{\partial {u^a_k}'}, \mathbf{n_i}'\right\rangle
= \psi_k(\mathbf{c_i})\, \hat{\mathbf{n}}_i^a$, the
full gradient evolution equation \eqref{eq:gradient_descent} yields:
\begin{align}
\frac{\partial E(u)}{\partial u^d_k} 
&= \underset{ij}{\sum} \underset{\mathbf{c}' \in \Gamma_{ij}}{\sum}
\left[ \Delta^2_i (\mathbf{f}' ) - \Delta^2_j (\mathbf{f}')\right]
\psi_k(\mathbf{c_i})\, \hat{\mathbf{n}}_i^a \notag\\
&2\, \underset{x,y,z}{\sum} a_d u^a_{k,d}
-2\, \underset{x,y,z}{\sum} b_d \nabla u^a_{k,d},
\label{eq:gradient_final}
\end{align}

\subsubsection{Semi-implicit Euler-forward optimization}
It is necessary to discretize $t$ in order to obtain a numerical
implementation of the equation \eqref{eq:gradient_final}:
\begin{align}
\frac{u^{t+1}-u^{t}}{\tau} 
&= - \underset{ij}{\sum} \underset{\mathbf{c}' \in \Gamma_{ij}}{\sum}
\left[ \Delta^2_i (\mathbf{f}' ) - \Delta^2_j (\mathbf{f}')\right]
\psi_k(\mathbf{c_i})\, \hat{\mathbf{n}}_i^a \notag\\
&-2\, \underset{x,y,z}{\sum} a_d (u^a_{k,d})^{t+1}
+2\, \underset{x,y,z}{\sum} b_d \nabla (u^a_{k,d})^{t+1}.
\end{align}

The associated Euler-Lagrange equation is found as:
\begin{equation}
(\tau^{-1} +2\, a_d - 2\, b_d \nabla )\,(u^a_{k,d})^{t+1} = 
\tau^{-1} (u^a_{k,d})^{t} - \frac{\partial E_{data}(u)}{\partial u^d_k},
\end{equation}
that is a linear system that we translate into Fourier domain,
to obtain the next deformation field $(u^a_{k,d})^{t+1}$:
\begin{multline}
(u^a_{k,d})^{t+1} = \\
 \mathcal{F}^{-1} \left\lbrace
\frac{\mathcal{F}\lbrace \tau^{-1} (u^a_{k,d})^{t} - \underset{ij}{\sum} \underset{\mathbf{c}' \in \Gamma_{ij}}{\sum}
\left[ \Delta^2_i (\mathbf{f}' ) - \Delta^2_j (\mathbf{f}')\right] \rbrace}
     {\mathcal{F}\lbrace (\tau^{-1} +2\,a_d)\mathcal{I} - 2\,b_d \Delta ) \rbrace}
     \right\rbrace
\end{multline}

@@

@@

\subsubsection{Convergence}
\todo[inline]{discuss choice of $\tau$, Courant-Friedrichs-Lewy (CFL) condition, etc.}

\subsubsection{Efficient field densification}
As long as the dense deformation field is iteratively interpolated
from the same set of control points that define the $L-1$ contours,
it is possible to pre-cache all the $\psi_k(c_i)$ weights into a
sparse matrix for fast densification. As well, we achieve a 
diffeomorphic transform by
